---
title: "Long-read Sequencing Quality Control Report"
author: "`r Sys.info()[['user']]`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: yes
    code_folding: hide
params:
  rootDir: NULL
  LRPipelineDir: NULL
  LOGenDir: NULL
  ERCCDir: NULL
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", warning=FALSE, message=FALSE, cache = FALSE)

# List of required packages
required_packages <- c("data.table", "ggplot2", "dplyr", "stringr", 
                       "reshape2", "scales", "cowplot", "ggfortify", "DT", "kableExtra")

# Check if packages are installed, and install missing ones
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos='https://cran.uk.r-project.org')
  }
}

# load packages
suppressMessages({
  lapply(required_packages, library, character.only = TRUE)
})

```

```{r setup, include=FALSE}
## Working on dropbox
#rootDir <- "C:/Users/sl693/Dropbox/Scripts/LRPipeline/vignette/pilot/"
#LRPipelineDir <- "C:/Users/sl693/Dropbox/Scripts/LRPipeline/"
#manifest <- read.csv(paste0(rootDir, "/0_metadata/manifestQC.csv"))
#ERCCDir <- "C:/Users/sl693/Dropbox/Scripts/LRPipeline/vignette/pilot/ERCC"
#LOGenDir <- "C:/Users/sl693/Dropbox/Scripts/LOGen/"

# rootDir = path to WKD_dir (root directory) of results from running pipeline
# LRPipelineDir = path to LRPipeline scripts directory 
# LOGenDir = path to Logen root directory
# manifest = path to manifest file
# ERCCDir = path to ERCC root directory if applicable
rootDir <- paste0(params$rootDir,"/")
LRPipelineDir <- paste0(params$LRPipelineDir,"/")
LOGenDir <- paste0(params$LOGenDir,"/")
ERCCDir <- params$ERCCDir

## ---------------- LOGEN scripts

# summary
source(paste0(LOGenDir, "transcriptome_stats/summarise_classfiles.R"))
source(paste0(LOGenDir, "transcriptome_stats/short_fragments_summary.R"))

# sensitivity, saturation
source(paste0(LOGenDir, "aesthetics_basics_plots/pthemes.R"))
source(paste0(LOGenDir,"longread_QC/power_calculations.R"))
source(paste0(LOGenDir,"transcriptome_stats/sample_sensitivity.R"))
source("ERCC_functions.R")

```

```{r load_input, include=FALSE}

load(paste0(rootDir, "/QC_input.RData"))


if("bulk"%in% input$manifest$cellType | any(is.na(input$manifest$cellType))){
  message("Bulk data, no cell specificity")
  cellSpecific = FALSE
}else{
  cellSpecific = TRUE
}

# references
if(unique(input$manifest$species) == "mouse"){
  geneType <- read.csv(paste0(LOGenDir,"0_utils/references/gencode.vM22.annotation.geneannotation_corrected.csv"), header = T)
}else{
  geneType <- read.csv(paste0(LOGenDir,"0_utils/references/gencode.v40.annotation.geneannotation_corrected.csv"), header = T)
}

numSample=length(unique(input$manifest$sample))
if(numSample == 1){
  multiplex = FALSE
}else{
  multiplex = TRUE
}
```

```{r load_ERCC, include=FALSE}

## ---------------- ERCC 

if(is.null(ERCCDir)){
  ERCC_added = FALSE
}else{
  ERCC_added = TRUE
  ERCCDir <- paste0(ERCCDir, "/")
}

# ERCC files if true
if(isTRUE(ERCC_added)){
  utilsDir <- paste0(LRPipelineDir, "utils/")
  ercc_calculation <- read.csv(paste0(utilsDir, "ERCC_calculations.csv"))
  ercc_stats <- read.csv(paste0(utilsDir, "ERCC_Stats.csv"))
  ercc_calculation <- merge(ercc_calculation, ercc_stats, by.x = "ERCC_ID", by.y = "ERCC.ID")

  file_path <- paste0(ERCCDir, "/", list.files(ERCCDir, pattern = "RulesFilter_result_classification\\.txt$", recursive = TRUE))
  message("Reading in ERCC file", file_path)
  ERCC_classFile <- fread(file_path, data.table = FALSE)
  ERCC_classFile <- ERCC_classFile %>% filter(filter_result == "Isoform") 
  monoexonic <- ERCC_classFile[ERCC_classFile$subcategory == "mono-exon" & ERCC_classFile$structural_category != "full-splice_match","isoform"]
  ERCC_classFile <- ERCC_classFile %>% filter(!isoform %in% monoexonic)
  ERCC_demux_file <- list.files(ERCCDir, pattern = "demux_fl_count\\.csv$", recursive = TRUE, full.names = TRUE)
  ERCC_demux <- fread(ERCC_demux_file, data.table = F)
}

```

# Study Information

This report documents the internal quality control (QC) process of long-read sequencing data generated at the University of Exeter Medical School for the following study:

**Title:** `r unique(input$manifest$studyName)`

**Date of QC:** `r format(Sys.Date(), format="%d %B %Y")`

**Species:** `r  unique(input$manifest$species)` 

**Number of samples:** `r  length(unique(input$manifest$sample))` 

**Number of individuals:** `r  length(unique(input$manifest$individual))` 

**Number of flow cells/batch:** `r  length(unique(input$manifest$batch))` 

**ERCC added:** `r ERCC_added`

**Sample tissue:** `r  unique(input$manifest$tissue)` 

**Project data path:** `r rootDir`

## Samples

Below is the full list of samples imported.

```{r manifest, fig.width = 12, fig.height = 6, echo = FALSE}
# manifest
datatable(input$manifest %>% select(-studyName),
          caption = "Phenotype information",
          options = list(
            pageLength = 5,
            autoWidth = TRUE  # Adjust column widths automatically
          ))

```

# QC: Sequencing quality  

A series of quality control (QC) sequencing metrics are provided below to aid identification of any patterns behind sample failures, artifacts that should be excluded and/or covariates that may be included in analysis. 

## Sequencing depth

For any given sequencing run (ONT flow cell) where multiplexing is performed, the number of reads obtained per sample post de-multiplexing should ideally be consistent. N50 is defined as the length of the shortest transcript at 50% of the total pool, and should be about ~1.5kb. Any significant variation in sample reads or small N50 are indicative of poor RNA quality. 

```{r demuxStats, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: bar plot shows the number of full-length (FL) reads obtained for each individual after sequencing (pre-alignment)."}

# table
knitr::kable(input$demuxStats[, c("sample", "num_seqs", "sum_len", "N50")], 
             col.names = c('sample', 'Number of reads', 'Number of bases', 'N50'),
             caption = "Table: summary of demultiplexing statistics for each sample, including the number of reads, the total number of bases, and the N50 value.") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))


# plot
if(isTRUE(cellSpecific)){
  p <- ggplot(input$demuxStats, aes(x = individual, y = num_seqs, fill = cellType)) 
}else{
  p <- ggplot(input$demuxStats, aes(x = individual, y = num_seqs)) 
}

p <- p +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_comma()) + theme_classic() +
  labs(x = "Individual", y = "Number of FL reads")

p
```

## Alignment rate

Post alignment to the reference genome, we are looking for a near-100% alignment identity (proportion of identical bases to the reference transcript) and a near-100% alignment coverage (proportion of bases matching to the reference transcript). While the alignment identity is an inherent measure of the sequencing platform (i.e. basecalling nanopore accuracy), the alignment coverage can provide insights into the sequencing library quality such that a poor alignment coverage could be suggestive of RNA degradation. 

```{r alignedStats, fig.width = 12, fig.height = 6, echo = FALSE}

# table
alignmentStats <- input$mappedStats %>%
  group_by(sample) %>%
  summarise(
    num_reads = n(),  
    mean_aligned_fraction = mean(aligned_fraction, na.rm = TRUE),
    mean_identity_percentage = mean(identity_percentage, na.rm = TRUE)
  )

knitr::kable(alignmentStats, col.names = c("Sample","number of reads", "aligned coverage", "aligned identity"),
             caption = "Table: summary of alignment statistics of the number of reads alignmed, the mean aligned coverage and identity of #those reads") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

For a transcriptome sequencing library, we are looking for a normal distribution around 1.5kb - 2kb; any peaks are indicative of RNA degradation or standards if added. 

```{r alignedPlot, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: distribution of the sample read length post alignment."}

ggplot(input$mappedStats, aes(x = sample, fill = sample, y = read_length)) +
  geom_violin() + theme_classic() +
  labs(y = "Read Length (bp)", x = "Sample") +
  theme(legend.position="none") + scale_y_log10() 

```

## Genome coverage

The genome coverage would depend if sequencing was performed on a whole transcriptome level or at a targeted level. 

```{r chromStats, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedrightFigure: number of reads aligned to each chromosome."}
dat <- input$mappedStats %>% group_by(sample, target_sequence) %>% tally() %>% 
  filter(!grepl("_", target_sequence)) %>% as.data.frame()

ggplot(dat, aes(x = as.factor(target_sequence), y = n)) + 
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(colour = sample)) +
  theme_classic() +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Chromosome", y = "Number of reads") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Standards

`r if (ERCC_added) {
  "
  The below metrics and plots relate to the detection of added ERCC. These are known molecules that are added to each sample during library preparation before multiplexing, and can be used to infer the library sensitivity and validity of the bioinformatics pipeline. There are 92 ERCC molecules of varying known amounts and length.
  "
} else {
  "No ERCC molecules were added to this sequencing run; QC unavailable."
}`


```{r ERCC, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedrightFigure: number of reads aligned to each chromosome.", eval = ERCC_added}
ERCC_alignedStats <- read_ERCC_alignedStats(paste0(ERCCDir,"3_minimap"))
detectedNumERCC <- length(unique(ERCC_alignedStats$target_sequence))

# tally the number of ERCC
numDetected <- ERCC_alignedStats %>% group_by(sample) %>% tally(name = "number")

# merge back with original number of reads per sample
numDetected <- merge(input$demuxStats[,c("sample","num_seqs")], numDetected, by = "sample") %>% 
  mutate(perc = number/num_seqs * 100)

numDetectedPerSample <- ERCC_alignedStats %>% group_by(sample) %>% summarise(unique_count = n_distinct(target_sequence))
numDetected <- merge(numDetected, numDetectedPerSample, by = "sample")

knitr::kable(numDetected, 
             col.names = c('Sample', '# demultiplexed reads', '# ERCC reads', "ERCC reads (%)", "# ERCC molecules"), 
             caption = "Number (#) and proportion (%) of ERCC reads per sample") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))

ERCC_detected <- detect_ERCC(ERCC_alignedStats)
```

`r if (ERCC_added) {
  "
  We expect that the ERCC detected would depend on the amount and not on the length i.e. ERCC with the known higher amounts are more likely to be sequenced, whereas longer ERCC are as likely to be sequenced as the shorter ERCC molecules. We also expect this phenomenon to be consistent across all the samples.
  "
}`


```{r amount, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: box-plot of the ERCC detected by amount.", eval = ERCC_added}
p <- plot_amount_length(ERCC_detected)
p[[1]]
```
```{r length, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: box-plot of the ERCC detected by length.", eval = ERCC_added}
p[[2]]
```

`r if (ERCC_added) {
  "### Detection accuracy\n#### Number of isoforms collapsed\n\nFollowing a similar bioinformatics pipeline, aligned ERCC reads are collapsed into unique isoforms and characterised using *SQANTI*. In theory, we should only detect one *isoform* per ERCC molecule. However, in practice, we may detect multiple *isoforms* annotated to the same ERCC molecule due to degradation and shearing during the library preparation."
}`


```{r redundancy1, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: The number of unique isoforms detected per ERCC molecule, coloured by sqanti structural categories", eval = ERCC_added}
pRedundant <- plot_numIsoforms(ERCC_classFile)
pRedundant[[1]]
```

`r if (ERCC_added) {
  "
  Notably, the number of *unique isoforms* detected per ERCC molecule is depending on the initial known amounts i.e. the higher the amount of that specific ERCC molecule, the more likely it is to be PCR amplified and sequenced.
  "
}`


```{r redundancy2, fig.width = 12, fig.height = 6, echo = FALSE, warning = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: The number of unique isoforms detected per ERCC molecule, coloured by sqanti structural categories", eval = ERCC_added}
pRedundant[[2]]
```

`r if (ERCC_added) {
  "
  Despite the significant number of *unique isoforms*, there should only be one or two **dominant** isoform per ERCC molecule whereby the remaining isoforms are insignificant - i.e. a *minor* isoform, defined by less than 5% of the total proportion.
  
  Below are plots of the *isoform* landscape of each ERCC molecule per sample. The number in the grey bar refers to the number of minor ERCC molecules, and the individual lines dissecting the bars refer to the proportion of the major isoform.
  "
}`


```{r usage, fig.width = 12, fig.height = 6, echo = FALSE, warning = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: bar plots of the isoform usage across each ERCC molecule", eval = ERCC_added}

pUsage <- plot_usage_persample(ERCC_classFile, ERCC_demux)
pUsage$plots[[1]]

```

```{r usageTable, fig.width = 12, fig.height = 6, echo = FALSE, warning = FALSE, eval = ERCC_added}

datatable(tabulate_isoformFraction(pUsage),
          caption = "Table of the number of dominant isoforms detected per ERCC per sample and it's corresponding proportion",
          filter = "top",     # Add column-specific filters
          options = list(
            autoWidth = TRUE,  # Adjust column widths automatically,
            scrollY = "auto",     # Set vertical scroll height
            paging = TRUE           # Enable pagination
          ))
```


# QC: Sample identity

A series of quality control (QC) sequencing metrics are provided below to ensure that the correct samples are assigned to the corresponding barcode/flow cell.

## Sex prediction

We can use the reads aligned to the X and Y chromosome to infer the sex of the samples, by calculating a fold change of the number of reads in the corresponding X and Y chromosome relative to the total reads aligned to the allosomes (X and Y together). 
- For females, we would expect a near 1 fold-change on the X chromosome and a near 0 fold change on the Y chromosome.  
- For males, we would expect a near 0.5 fold change on the X chromosome and Y chromosome.   

```{r sexCheck, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: number of reads aligned to X and Y chromosome over total reads aligned to X and Y chromosome"}

# X chromosome, Y chromsoome, autosomal reads
xReads <- input$filteredAlignment %>% filter(target_sequence %in% c("chrX")) %>% group_by(sample) %>% tally(name = "chrxReads")
yReads <- input$filteredAlignment %>% filter(target_sequence %in% c("chrY")) %>% group_by(sample) %>% tally(name = "chrYReads")
autoReads <- input$filteredAlignment %>% filter(!target_sequence %in% c("chrX","chrY")) %>% group_by(sample) %>% tally(name = "autoReads")

# fold cange of chrX over all reads, chrY over all reads
sexCheckReads <- Reduce(function(x, y) merge(x, y, by = "sample", all = TRUE), list(xReads, yReads, autoReads)) %>%
  replace(is.na(.), 0) %>%
  mutate(FCX = chrxReads/(chrxReads + chrYReads), FCY = chrYReads/(chrxReads + chrYReads))
sexCheckReads <- merge(sexCheckReads, input$manifest, by = "sample")

# plot
ggplot(sexCheckReads, aes(x = FCX, y = FCY, colour = sex)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(x = "X ratio", y = "Y ratio")
```

# QC: Data quality

## RT artifacts

As part of the final bioinformatics pipieline, reads are collopsed into transcripts and are filtered by *SQANTI* if they are classified as artifacts in the form of:  
- intra-priming, defined by high percentage of A downstream of transcription termination site  
- reverse-transcription switching  

```{r reasonFiltered, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: box-plot of the number of reads classified as artifacts from reverse-transcription step, and subsequently removed"}

# isolate the reads that are corresponding to the artifacts
artifacts <- input$preFilteredClassFile[input$preFilteredClassFile$filter_result == "Artifact","isoform"]
passedReads <- input$preFilteredClassFile[input$preFilteredClassFile$filter_result == "Isoform","isoform"]
artifactDemux <- input$demux[input$demux$id %in% artifacts,] 
artifactDemux <- merge(artifactDemux, input$ReasonFiltered[,c("isoform","reasons")], by.x = "id", by.y = "isoform")
artifactSummary <- artifactDemux %>% 
  select(-id) %>%
  group_by(reasons) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# stepwise stats filtering
stage1Stats <- input$demuxStats[,c("sample","num_seqs")]
stage2Stats <- as.data.frame(input$mappedStats %>% group_by(sample) %>% tally(name = "aligned_seqs"))
stage3Stats <- as.data.frame(input$filteredAlignment %>% group_by(sample) %>% tally(name = "corrected_seqs"))
stage4Stats <- as.data.frame(input$demux %>% select(-id) %>% apply(.,2,sum))       
colnames(stage4Stats) <- "collapsedReads"
passedReads <- input$preFilteredClassFile[input$preFilteredClassFile$filter_result == "Isoform","isoform"] 
stage5Stats <- as.data.frame(input$demux %>% filter(id %in% passedReads) %>% select(-id) %>% apply(.,2,sum))              
colnames(stage5Stats) <- "finalReads"
# demux file to keep only the finalised isoforms
stage6Stats <- as.data.frame(input$demuxFiltered %>% apply(.,2,sum)) 
colnames(stage6Stats) <- "monoFilteredReads"

# filter demux file
# Define a custom function to count non-zero elements in a row
count_non_zero <- function(row) {
  return(sum(row != 0))
}
demuxFiltered <- input$demuxFiltered
nreads <- apply(input$demuxFiltered,1,sum)
nsample <- apply(input$demuxFiltered, 1, count_non_zero)
demuxFiltered$nreads <- nreads
demuxFiltered$nsample <- nsample
demuxFiltered <- demuxFiltered %>% filter(nreads >= 2, nsample >= 2)

stage7Stats <- as.data.frame(demuxFiltered %>% select(-nreads, -nsample) %>% apply(.,2,sum)) 
colnames(stage7Stats) <- "expressionFilteredReads"
stage4Stats <- merge(input$manifest[,c("ID","sample")], stage4Stats, by.x = "ID", by.y = 0)
stage5Stats <- merge(input$manifest[,c("ID","sample")], stage5Stats, by.x = "ID", by.y = 0)
stage6Stats <- merge(input$manifest[,c("ID","sample")], stage6Stats, by.x = "ID", by.y = 0)
stage7Stats <- merge(input$manifest[,c("ID","sample")], stage7Stats, by.x = "ID", by.y = 0)

fullStats <- Reduce(function(x, y) merge(x, y, by = "sample", all = TRUE), 
                    list(stage1Stats, stage2Stats, stage3Stats, stage4Stats, stage5Stats, stage6Stats, stage7Stats)) 

fullStats <- fullStats %>% select(-contains("ID")) %>% 
  mutate(percCorrectedAligned = corrected_seqs/num_seqs * 100, 
         percCollapsed = collapsedReads/num_seqs * 100,
         percFinal = finalReads/num_seqs * 100,
         percMonofiltered = monoFilteredReads/num_seqs * 100,
         percExpressionFiltered = expressionFilteredReads/num_seqs * 100) 

fullStats <- merge(fullStats, input$manifest, by = "sample")

artifactSummary$reasons <- gsub("High perc_A_downstream_TTS", "intra-priming", artifactSummary$reasons)
artifactSummary$reasons <- gsub("Out RTS_stage", "RT switching", artifactSummary$reasons)
reshape2::melt(artifactSummary, value.name = "removedFL", variable.name = "ID") %>%
  merge(., fullStats, by = "ID") %>% 
  mutate(percRemoved = removedFL/num_seqs * 100) %>% 
  ggplot(., aes(x = reasons, y = percRemoved)) + 
  geom_boxplot(outliers = FALSE) +
  geom_jitter(aes(colour = sample)) + theme_classic() +
  labs(x = "Reasons filtered", y = "Percentage of reads removed")
```

## Mono-exonic isoforms

Shown below is a table of the number of mono-exonic transcripts (transcripts with 1 exon) and corresponding full-length reads detected in the dataset (across all samples) prior to SQANTI filtering (see section above). 

::: note
Further filtering refers to keeping only mono-exonic transcripts that are classified as FSM, ISM, NIC, NNC, Antisense and Fusion
:::

```{r monoExonic}
monoExonicSummary <- monoExonic_summary(input$preFilteredClassFile, input$demux)

knitr::kable(monoExonicSummary, 
             col.names = c(''),
             caption = "Table: Summary of monoexonic transcripts"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

Below is a plot of the length of mono-exonic transcripts in the dataset classified as artifact and isoform by SQANTI filtering.

```{r monoExonicLength, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: Violin plot of the length of mono-exonic transcripts by structural category"}
input$preFilteredClassFile %>% filter(exons == 1) %>%
  ggplot(., aes(x = structural_category, y = length, fill = filter_result)) +
  geom_violin() +scale_y_log10() + theme_classic() +
  labs(x = "structural_category", y = "length (bp)") 

```
## Short isoforms

A good quality metric of the library prepartion/sample/sequencing run is the proportion of short isoforms. These short isoforms could either be real and biologically meaningful (i.e. short regulatory RNA) or artifacts from library preparation, which are typically presented as short fragments randomly distributed across the gene/genome i.e intergenic/genic intron/genic genomic. 

Tabulated below are the number of short transcripts detected in the dataset after applying a length threshold (200bp, 500bp, 750bp), and the number kept:  
- after applying SQANTI filter (removing intrapriming and RT switch artifacts)  
- further removing transcripts identified as mono-exonic and classified as either mono-exonic intergenic, genic intronic or genic genomic  
- keeping transcripts that are detected with 2 reads across any 2 samples.

::: note
All the transcripts that are kept under the 200bp threshold will also be kept in the 500bp threshold, and likewise those in the 500bp would also be kept in the 750bp. The purpose of the table is to identify the length at which there is a dramatic increase of short transcripts detected, even when the final proportion of "TRUE meaningful transcripts" kept is low. 
:::

```{r shortIsoforms}

# different length threshold to test
lengths_to_test <- c(200, 500, 750)

# create column nreads and nsamples in demux file
demux <- input$demux
row.names(demux) <- demux$id
demux <- demux %>% select(-id)
nreads <- apply(demux,1,sum)
nsample <- apply(demux, 1, count_non_zero)
demux$nreads <- nreads
demux$nsample <- nsample

# generate subsetted classification file by length
df <- lapply(lengths_to_test, function(x) subset_class_by_length(x, input$preFilteredClassFile, demux, input$ReasonFiltered, geneType))
names(df) <- lengths_to_test

# generate summary stats of the subsetted classification file by length
dfStats <- lapply(df, function(x) summarise_subsetted_classLength(x, input$preFilteredClassFile))
outputStats <- t(do.call(rbind, dfStats))
colnames(outputStats) <- lengths_to_test

knitr::kable(outputStats, caption = "Table: Summary of short transcripts") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))

```

Below is a table of the full-length reads of short transcripts

```{r shortIsoformsReads}

dfStats <- lapply(df, function(x) summarise_subsetted_classLength_read(x, demux))
outputStats <- t(do.call(rbind, dfStats))
colnames(outputStats) <- lengths_to_test

knitr::kable(outputStats, caption = "Table: Summary of short transcripts at the read level") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))

```

Below is a plot of the proportion of short transcripts (with the length threshold of 200bp, 500bp, 750bp) that are discarded during SQANTI filtering either due to RT switching (Out RT stage), intrapriming (High perc_A_downstream_TTS) or lack of support of non-canonical junctions. 

```{r shortIsoformsFiltered, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: proportion of short transcripts discarded during SQANTI filtering"}
SQANTIFilteredReasonDf <- lapply(df, function(x) subset_class_length_SQFiltered(x))
SQANTIFilteredReasonDf <- rbindlist(SQANTIFilteredReasonDf, idcol = "lengthCutOff") 

ggplot(SQANTIFilteredReasonDf, aes(y = reasons, x = perc, fill = structural_category)) + geom_bar(stat = "identity") +
  facet_grid(lengthCutOff~.) +
  labs(x = "Percentage of reads discarded", y = "Reasons for SQANTI filter") +
  theme_classic()

```

## Expression threshold

To reduce false negatives and increase confidence of identifying "true" isoforms, we can apply an expression filter threshold i.e. an isoform is kept only if it has more than 2 reads across any 2 samples.

The below plot shows the proportion of reads retained after applying a step-wise threshold. Each line corresponds to the minimum number of reads and the x-axis refers to the minimum number of samples. 

```{r sensitivity, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: Line plot of the number of reads retained post SQANTI filtering after applying an expression threshold"}

sensitivity <- input$demuxFiltered 
colnames(sensitivity) <- paste0("FL.", colnames(sensitivity))
sensitivtyOutput <- no_of_isoforms_sample(sensitivity)
sensitivtyOutput[[3]]$cate_sumFL <- categories(as.numeric(as.character(sensitivtyOutput[[3]]$sum_FL)))

dat2 <- sensitivtyOutput[[3]] %>%
  group_by(num_samples, cate_sumFL) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = count / sum(count)) %>%
  mutate(cate_sumFL = factor(cate_sumFL, levels = c(seq(1:9),"10","10-20","21-30","31-40","41-50","> 50")))

ggplot(dat2, aes(x = as.factor(num_samples), y = proportion, colour = cate_sumFL, group = cate_sumFL)) +
  geom_point() +
  geom_line() + scale_y_continuous(labels = perc_lab)  + 
  labs(x = "Minimum number of samples", y = "Reads (%)",
       colour = "Minimum number of reads") +
  theme_classic() 
  
```

# QC: Sample reproducibility

We anticipate that while the number of reads may differ across the samples, the extent of reads retained after each processing step will be consistent. Any deviations will suggest a sample/batch failure. 

The plot displays the proportion of reads in each sample **retained** after the following steps:  
    1. Alignment and correction  
    2. Collapsing reads to transcripts  
    3. Filtering reads for RT artifacts  
    4. Removing mono-exonic non-FSM reads  
    5. Applying a minimum expression threshold  

```{r fullStats, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: proportion of reads retained after each processing/filtering stage"}
fullStatsPlot <- fullStats %>% select(sample, contains("perc")) %>% 
  reshape2::melt(variable.name = "stage", value.name = "perc") %>% 
  filter(perc != "NA") 

xaxislevelsF1 <- c("percCorrectedAligned","percCollapsed","percFinal","percMonofiltered", "percExpressionFiltered")
xaxislabelsF1 <- c("Alignment", "Collapse", "SQANTI filtering", "monoexonic filtering", "Expression filtering")
fullStatsPlot$stage = factor(fullStatsPlot$stage,
                             labels = xaxislabelsF1, 
                             levels = xaxislevelsF1,
                             ordered=TRUE)

ggplot(fullStatsPlot , aes(x = stage, y = perc, group = sample, colour = sample)) +
  geom_point() +
  geom_line() +
  theme_classic() + 
  labs(y = "Percentage of reads", x = "Stage") +
  ylim(0,100)
```

# QC: Biological validity

The following sections assess whether the samples sequenced reflect biological phenomena and if there are any covariates that should be accounted in downstream analysis. 



## Normalisation

```{r normalisation, fig.width = 12, fig.height = 6, echo = FALSE}

# calculate TPM 
# https://github.com/ConesaLab/SQANTI3/wiki/Running-SQANTI3-Quality-Control
# raw FL read for transcript in sample 1/total FL transcript in sample 1 * 10^6
demuxCollapsed <- input$demux %>% filter(id %in% input$preFilteredClassFile$isoform)
row.names(demuxCollapsed) <- input$demux$id
demuxTPM <- apply(demuxCollapsed %>% select(-id),2,function(x){x/sum(x) * 1000000})
demuxFilteredTPM <- apply(input$demuxFiltered,2,function(x){x/sum(x) * 1000000})

```

## Gene expression

We expect the gene expression of house keeping genes to be consistent across all samples, and differences in cell-specific genes for sorted populations.

```{r geneCellExpressions, fig.width = 12, fig.height = 6, echo = FALSE}
plot_geneExp <- function(gene){
  gclass_file <- input$preFilteredClassFile %>% filter(associated_gene == gene) %>% select(isoform, associated_gene)
  
  geneExp <- merge(gclass_file, demuxTPM, by.x = "isoform", by.y = 0) %>% 
    select(-isoform,-associated_gene) %>%
    apply(., 2, sum) %>% 
    reshape2::melt(value.name = "exp") 
  
  geneExp <- merge(input$manifest, geneExp, by.x = "ID", by.y = 0)
  geneExp <<- geneExp  
  
  if("cellType" %in% colnames(input$manifest)){
  p <- ggplot(geneExp, aes(x = cellType, y = exp)) + geom_boxplot(outliers = FALSE) + 
    geom_jitter(size = 3, aes(colour = individual)) + 
    theme_classic() + 
    labs(x = "Sorted populations", y = "Expression", title = gene)
  }
  
  return(p)
}


if(unique(input$manifest$species) == "mouse"){
  housekeepingGene <- c("Gapdh")
  plot_geneExp("Gapdh")
}else{
  housekeepingGene <- c("GAPDH")
  plot_geneExp("GAPDH")
}

```

### Cell-specific genes

`r if (cellSpecific) {
  "
  The below metrics and plots relate to the gene expression profiles of cell-specific markers. 
  "
} else {
  "Experiment was performed on bulk data - cell specificity cannot be determined"
}`


```{r neuronalExpression, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: neuronal-specific gene expressions", eval = cellSpecific}

if(unique(input$manifest$species) == "mouse"){
  NeuronalGenes <- c("Cabp1","Prrg3","Rbfox3","Eno2")
}else{
  NeuronalGenes <- c("CABP1","PRRG3","RBFOX3","ENO2")
}

NeuonralGenePlots <- lapply(NeuronalGenes, function(gene) plot_geneExp(gene))
plot_grid(plotlist = NeuonralGenePlots)

```

```{r oligoCellExpressions, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: oligodendrocyte-specific gene expressions", eval = cellSpecific}

if(unique(input$manifest$species) == "mouse"){
  OligoGenes <-  c("Sox10", "Opalin","St18","Olig1")
}else{
  OligoGenes <-  c("SOX10", "OPALIN","ST18","OLIG1")
}

OligoGenePlots <- lapply(OligoGenes, function(gene) plot_geneExp(gene))
plot_grid(plotlist = OligoGenePlots)

```

# QC: Saturation 

`r if (multiplex) {
  "
  
  Below are rarefaction curves of the coverage of RNA isoform diversity by SQANTI's structural category with added samples. We anticipate a plateau of the number of new transcripts added with each sequential sample. 
  Note:  The more abundant transcripts are always detected first, so FSM transcripts will also be the first to be detected compared to novel transcripts. 
  "
} else {
  "Saturation curve not available for one sample."
}`



```{r saturation_plot, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: Cumulative number of transcripts with added sample", eval = multiplex}
class_file <- merge(input$class_file, input$demuxFiltered, by.x = "isoform", by.y = 0)
plot_cumulative_added_sample(class_file, input$demuxFiltered)
```

# Descriptives

## Transcriptome numbers  

Below is a basic summary of the number of transcripts and genes found in the transcriptome of all the samples.

```{r summary} 
class_file <- merge(input$class_file, input$demuxFiltered, by.x = "isoform", by.y = 0)
class_file_annoGenes <- class_file[!grepl("novel", class_file$associated_gene),]

summary <- descriptives_summary(class_file, class_file_annoGenes)
knitr::kable(summary, 
             col.names = c(''),
             caption = "Table: Summary of transcriptome"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

## Type of transcripts

```{r geneType, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: number of transcripts by gene type"}
class_file <- merge(class_file, distinct(geneType[,c("GeneSymbol","Class")]), by.x = "associated_gene", by.y = "GeneSymbol", all.x = T)

class_file %>% group_by(Class) %>% tally() %>% 
  ggplot(., aes(y = reorder(Class,n), x = n)) + geom_bar(stat = "identity") +
  scale_x_log10(labels = label_comma()) +
  labs(x = "number of transcripts", y = "Class") + 
  theme_classic() 

```

## Top-ranked genes

```{r geneExpression}

dat <- class_file %>% 
  select(associated_gene, colnames(input$demuxFiltered)) %>%
  group_by(associated_gene) %>% 
  summarise(across(everything(), sum)) %>% 
  tibble::column_to_rownames(., var = "associated_gene")

meanGeneExp <- as.data.frame(apply(dat, 1, mean))
colnames(meanGeneExp) <- "Mean_gene_expression"
meanGeneExp <- meanGeneExp %>% tibble::rownames_to_column(., var = "associated_gene") 
meanGeneExp <- merge(meanGeneExp, distinct(geneType[,c("GeneSymbol","Class")]), by.x = "associated_gene", by.y = "GeneSymbol", all.x = T) %>% arrange(-Mean_gene_expression)

datatable(meanGeneExp[1:100,],
          caption = "Top-ranked 100 genes by expression",
          options = list(
            pageLength = 5,
            autoWidth = TRUE  # Adjust column widths automatically
          ))

```

Session
```{r sessionInfo}
sessionInfo()
```
