---
title: "Long-read Sequencing Quality Control Report"
author: "`r Sys.info()[['user']]`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: yes
    code_folding: hide
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", warning=FALSE, message=FALSE, cache = FALSE)
```

```{r setup, include=FALSE}
args<-commandArgs(trailingOnly = TRUE)
args[1] <- "C:/Users/sl693/Dropbox/Scripts/LRPipeline/vignette/pilot"
rootDir <- args[1]

suppressMessages(library("data.table"))
suppressMessages(library("ggplot2"))
suppressMessages(library("dplyr"))
suppressMessages(library("stringr"))
suppressMessages(library("reshape2"))
suppressMessages(library("data.table"))
suppressMessages(library("scales"))
suppressMessages(library("cowplot"))
suppressMessages(library("ggfortify"))
suppressMessages(library("DT"))

studyName = "Human sorted nuclei dataset"
personSequenced = "Rosie Bamford"
manifest <- read.csv(paste0(rootDir, "/0_metadata/manifest.csv"))

# utils ercc
utilsDir <- "C:/Users/sl693/Dropbox/Scripts/LRPipeline/ont_cDNA/utils/"
ercc_calculation <- read.csv(paste0(utilsDir, "ERCC_calculations.csv"))
ercc_stats <- read.csv(paste0(utilsDir, "ERCC_Stats.csv"))
ercc_calculation <- merge(ercc_calculation, ercc_stats, by.x = "ERCC_ID", by.y = "ERCC.ID")

# sensitivity
source("C:/Users/sl693/Dropbox/Scripts/LOGen/aesthetics_basics_plots/pthemes.R")
source("C:/Users/sl693/Dropbox/Scripts/LOGen/transcriptome_stats/sample_sensitivity.R")
source("ERCC_functions.R")

# ERCC files
ERCC_added = FALSE
if(isTRUE(ERCC_added)){
  ERCC_rootDir <- "C:/Users/sl693/Dropbox/Scripts/LRPipeline/vignette/pilot/ERCC"
  ERCC_classFile <- fread(paste0(ERCC_rootDir, "/AlessiaERCC_collapsed_RulesFilter_result_classification.txt"), data.table = FALSE)
  ERCC_classFile <- ERCC_classFile %>% filter(filter_result == "Isoform") 
  monoexonic <- ERCC_classFile[ERCC_classFile$subcategory == "mono-exon" & ERCC_classFile$structural_category != "full-splice_match","isoform"]
  ERCC_classFile <- ERCC_classFile %>% filter(!isoform %in% monoexonic)
  ERCC_demux <- fread(paste0(ERCC_rootDir, "/demux_fl_count.csv"), data.table = F)
}

```

# Study Information

This report documents the internal quality control (QC) process of long-read sequencing data generated at the University of Exeter Medical School for the following study:

**Title:** `r studyName`

**Date of QC:** `r format(Sys.Date(), format="%d %B %Y")`

**Species:** `r  unique(manifest$species)` 

**Number of samples:** `r  length(unique(manifest$sample))` 

**Number of individuals:** `r  length(unique(manifest$individual))` 

**Number of flow cells/batch:** `r  length(unique(manifest$batch))` 

**Sample tissue:** `r  unique(manifest$tissue)` 

**Samples sequenced by:** `r  personSequenced` 

**Project data path:** `r rootDir`

## Samples

Below is the full list of samples imported.

```{r manifest, fig.width = 12, fig.height = 6, echo = FALSE}
# manifest
datatable(manifest,
          caption = "Phenotype information",
          options = list(
            pageLength = 5,
            autoWidth = TRUE  # Adjust column widths automatically
          ))

```

# QC: Sequencing quality  

A series of quality control (QC) sequencing metrics are provided below to aid identification of any patterns behind sample failures, artifacts that should be excluded and/or covariates that may be included in analysis. 

## Sequencing depth

For any given sequencing run (ONT flow cell) where multiplexing is performed, the number of reads obtained per sample post de-multiplexing should ideally be consistent. N50 is defined as the length of the shortest transcript at 50% of the total pool, and should be about ~1.5kb. Any significant variation in sample reads or small N50 are indicative of poor RNA quality. 

```{r demuxStats, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: bar plot shows the number of full-length (FL) reads obtained for each individual after sequencing (pre-alignment)."}

# data-wrangle for readstats.txt for demultiplexed samples
demuxStats <- lapply(list.files(path = paste0(rootDir, "/1b_demultiplex_merged/"), pattern = "readstats", full = T), function(x) 
  fread(x, data.table = F))
demuxStats <- rbindlist(demuxStats, idcol = "sampleNum") %>% mutate(sample = word(file,c(1),sep=fixed("_")))
demuxStats$num_seqs <- as.numeric(gsub(",", "", demuxStats$num_seqs))
demuxStats <- merge(demuxStats, manifest, by = "sample") 

# table
knitr::kable(demuxStats[,c("sample","num_seqs","sum_len","N50")], 
             col.names = c('sample', 'Number of reads', 'Number of bases', 'N50'),
             caption = "Table: summary of demultiplexing statistics for each sample, including the number of reads, the total number of bases, and the N50 value."
)

# plot
if("cellType" %in% colnames(manifest)){
  p <- ggplot(demuxStats, aes(x = individual, y = num_seqs, fill = cellType)) 
}else{
  p <- ggplot(demuxStats, aes(x = individual, y = num_seqs)) 
}

p <- p +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_comma()) + theme_classic() +
  labs(x = "Individual", y = "Number of FL reads")

p
```

## Alignment rate

Post alignment to the reference genome, we are looking for a near-100% alignment identity (proportion of identical bases to the reference transcript) and a near-100% alignment coverage (proportion of bases matching to the reference transcript). While the alignment identity is an inherent measure of the sequencing platform (i.e. basecalling nanopore accuracy), the alignment coverage can provide insights into the sequencing library quality such that a poor alignment coverage could be suggestive of RNA degradation. 

```{r alignedStats, fig.width = 12, fig.height = 6, echo = FALSE}

# input of mapped stats 
nameAlignedStats <- c(
  "read_name",                # Name of the nanopore read
  "target_sequence",          # Name of the target sequence
  "start_position",           # Start position of the alignment on the target sequence
  "read_length",              # Length of the original nanopore read
  "aligned_length",           # Length of the aligned part of the nanopore read
  "aligned_fraction",         # Fraction of the aligned part of the nanopore read over the original length
  "matches",                  # Number of matched nucleotides in the alignment
  "identity_percentage",      # Fraction of matched nucleotides over the aligned length
  "strand",                   # Strand where the nanopore read aligns
  "mismatches",               # Number of mismatches in the alignment
  "insertions",               # Number of insertions in the alignment
  "deletions"                 # Number of deletions in the alignment
)
mappedStats <- lapply(list.files(path = paste0(rootDir, "/3_minimap/"), pattern = "mapped", full = T), function(x) 
  fread(x, col.names = nameAlignedStats, data.table = F))
names(mappedStats) <- list.files(path = paste0(rootDir, "/3_minimap/"), pattern = "mapped")
mappedStats <- rbindlist(mappedStats, idcol = "file") 
mappedStats$sample <- str_remove(mappedStats$file, "_merged_combined_mappedstats.txt")

# table
alignmentStats <- mappedStats %>%
  group_by(sample) %>%
  summarise(
    num_reads = n(),  
    mean_aligned_fraction = mean(aligned_fraction, na.rm = TRUE),
    mean_identity_percentage = mean(identity_percentage, na.rm = TRUE)
  )

knitr::kable(alignmentStats, col.names = c("Sample","number of reads", "aligned coverage", "aligned identity"),
             caption = "Table: summary of alignment statistics of the number of reads alignmed, the mean aligned coverage and identity of those reads") 
```

For a transcriptome sequencing library, we are looking for a normal distribution around 1.5kb - 2kb; any peaks are indicative of RNA degradation or standards if added. 

```{r alignedPlot, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: distribution of the sample read length post alignment."}

ggplot(mappedStats, aes(x = read_length, fill = sample)) +
  geom_density(alpha = 0.2) + theme_classic() +
  labs(x = "Read Length")

```

## Genome coverage

The genome coverage would depend if sequencing was performed on a whole transcriptome level or at a targeted level. 

```{r chromStats, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedrightFigure: number of reads aligned to each chromosome."}
dat <- mappedStats %>% group_by(sample, target_sequence) %>% tally() %>% 
  filter(!grepl("_", target_sequence)) %>% as.data.frame()

ggplot(dat, aes(x = as.factor(target_sequence), y = n)) + 
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(colour = sample)) +
  theme_classic() +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Chromosome", y = "Number of reads")
```

## Standards

`r if (ERCC_added) {
  "
  The below metrics and plots relate to the detection of added ERCC. These are known molecules that are added to each sample during library preparation before multiplexing, and can be used to infer the library sensitivity and validity of the bioinformatics pipeline. There are 92 ERCC molecules of varying known amounts and length.
  "
} else {
  "No ERCC molecules were added to this sequencing run; QC unavailable."
}`


```{r ERCC, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedrightFigure: number of reads aligned to each chromosome.", eval = ERCC_added}
ERCC_alignedStats <- read_ERCC_alignedStats(ERCC_rootDir)
detectedNumERCC <- length(unique(ERCC_alignedStats$target_sequence))

# tally the number of ERCC
numDetected <- ERCC_alignedStats %>% group_by(sample) %>% tally(name = "number")

# merge back with original number of reads per sample
numDetected <- merge(demuxStats[,c("sample","num_seqs")], numDetected, by = "sample") %>% 
  mutate(perc = number/num_seqs * 100)

numDetectedPerSample <- ERCC_alignedStats %>% group_by(sample) %>% summarise(unique_count = n_distinct(target_sequence))
numDetected <- merge(numDetected, numDetectedPerSample, by = "sample")

knitr::kable(numDetected, col.names = c('Sample', '# demultiplexed reads', '# ERCC reads', "ERCC reads (%)", "# ERCC molecules"), caption = "Number (#) and proportion (%) of ERCC reads per sample")

ERCC_detected <- detect_ERCC(ERCC_alignedStats)
```

`r if (ERCC_added) {
  "
  We expect that the ERCC detected would depend on the amount and not on the length i.e. ERCC with the known higher amounts are more likely to be sequenced, whereas longer ERCC are as likely to be sequenced as the shorter ERCC molecules. We also expect this phenomenon to be consistent across all the samples.

  ### Factors impacting detection
  "
}`


```{r amount, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: box-plot of the ERCC detected by amount.", eval = ERCC_added}
p <- plot_amount_length(ERCC_detected)
p[[1]]
```
```{r length, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: box-plot of the ERCC detected by length.", eval = ERCC_added}
p[[2]]
```

`r if (ERCC_added) {
  "### Detection accuracy\n#### Number of isoforms collapsed\n\nFollowing a similar bioinformatics pipeline, aligned ERCC reads are collapsed into unique isoforms and characterised using *SQANTI*. In theory, we should only detect one *isoform* per ERCC molecule. However, in practice, we may detect multiple *isoforms* annotated to the same ERCC molecule due to degradation and shearing during the library preparation."
}`


```{r redundancy1, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: The number of unique isoforms detected per ERCC molecule, coloured by sqanti structural categories", eval = ERCC_added}
pRedundant <- plot_numIsoforms(ERCC_classFile)
pRedundant[[1]]
```

`r if (ERCC_added) {
  "
  Notably, the number of *unique isoforms* detected per ERCC molecule is depending on the initial known amounts i.e. the higher the amount of that specific ERCC molecule, the more likely it is to be PCR amplified and sequenced.
  "
}`


```{r redundancy2, fig.width = 12, fig.height = 6, echo = FALSE, warning = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: The number of unique isoforms detected per ERCC molecule, coloured by sqanti structural categories", eval = ERCC_added}
pRedundant[[2]]
```

`r if (ERCC_added) {
  "
  #### Isoform proportions
  
  Despite the significant number of *unique isoforms*, there should only be one or two **dominant** isoform per ERCC molecule whereby the remaining isoforms are insignificant - i.e. a *minor* isoform, defined by less than 5% of the total proportion.
  
  Below are plots of the *isoform* landscape of each ERCC molecule per sample. The number in the grey bar refers to the number of minor ERCC molecules, and the individual lines dissecting the bars refer to the proportion of the major isoform.
  "
}`


```{r usage, fig.width = 12, fig.height = 6, echo = FALSE, warning = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: bar plots of the isoform usage across each ERCC molecule", eval = ERCC_added}

pUsage <- plot_usage_persample(ERCC_classFile, ERCC_demux)
pUsage$plots[[1]]

```

```{r usageTable, fig.width = 12, fig.height = 6, echo = FALSE, warning = FALSE, eval = ERCC_added}

datatable(tabulate_isoformFraction(pUsage),
          caption = "Table of the number of dominant isoforms detected per ERCC per sample and it's corresponding proportion",
          filter = "top",     # Add column-specific filters
          options = list(
            autoWidth = TRUE,  # Adjust column widths automatically,
            scrollY = "auto",     # Set vertical scroll height
            paging = TRUE           # Enable pagination
          ))
```


# QC: Sample identity

A series of quality control (QC) sequencing metrics are provided below to ensure that the correct samples are assigned to the corresponding barcode/flow cell.

## Sex prediction

We can use the reads aligned to the X and Y chromosome to infer the sex of the samples, by calculating a fold change of the number of reads in the corresponding X and Y chromosome relative to the total reads aligned to the allosomes (X and Y together). 
- For females, we would expect a near 1 fold-change on the X chromosome and a near 0 fold change on the Y chromosome.  
- For males, we would expect a near 0.5 fold change on the X chromosome and Y chromosome.   

```{r sexCheck, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: number of reads aligned to X and Y chromosome over total reads aligned to X and Y chromosome"}

# data-wrangle of final filtered reads
filteredAlignment <- lapply(list.files(path = paste0(rootDir, "/5_cupcake/5_align/PAF"), pattern = "mappedstats", full = T), function(x) 
  fread(x, data.table = F, col.names = nameAlignedStats))
names(filteredAlignment) <- list.files(path = paste0(rootDir, "/5_cupcake/5_align/PAF"), pattern = "mappedstats")
filteredAlignment <- rbindlist(filteredAlignment, idcol = "file") 
filteredAlignment$sample <- str_remove(filteredAlignment$file, "_mapped_mappedstats.txt")

# X chromosome, Y chromsoome, autosomal reads
xReads <- filteredAlignment  %>% filter(target_sequence %in% c("chrX")) %>% group_by(sample) %>% tally(name = "chrxReads")
yReads <- filteredAlignment  %>% filter(target_sequence %in% c("chrY")) %>% group_by(sample) %>% tally(name = "chrYReads")
autoReads <- filteredAlignment  %>% filter(!target_sequence %in% c("chrX","chrY")) %>% group_by(sample) %>% tally(name = "autoReads")

# fold cange of chrX over all reads, chrY over all reads
sexCheckReads <- Reduce(function(x, y) merge(x, y, by = "sample", all = TRUE), list(xReads, yReads, autoReads)) %>%
  mutate(FCX = chrxReads/(chrxReads + chrYReads), FCY = chrYReads/(chrxReads + chrYReads))
sexCheckReads <- merge(sexCheckReads, manifest, by = "sample")

# plot
ggplot(sexCheckReads, aes(x = FCX, y = FCY, colour = sex)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(x = "X ratio", y = "Y ratio")
```

# QC: Data quality

## RT artifacts

As part of the final bioinformatics pipieline, reads are collopsed into transcripts and are filtered by *SQANTI* if they are classified as artifacts in the form of:  
- intra-priming, defined by high percentage of A downstream of transcription termination site  
- reverse-transcription switching  

```{r reasonFiltered, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: box-plot of the number of reads classified as artifacts from reverse-transcription step, and subsequently removed"}

# classifiction file with all isoforms and artifacts
preFilteredClassFileName <- list.files(path = paste0(rootDir,"/5_cupcake/7_sqanti3"), pattern = "RulesFilter_result_classification", full = T)
preFilteredClassFile <- fread(preFilteredClassFileName, data.table = F)
xaxislevelsF1 <- c("full-splice_match","incomplete-splice_match","novel_in_catalog","novel_not_in_catalog","genic","antisense","fusion","intergenic","genic_intron")
xaxislabelsF1 <- c("FSM", "ISM", "NIC", "NNC", "Genic_Genomic",  "Antisense", "Fusion","Intergenic", "Genic_Intron")
preFilteredClassFile$structural_category = factor(preFilteredClassFile$structural_category,
                                        labels = xaxislabelsF1, 
                                        levels = xaxislevelsF1,
                                        ordered=TRUE)

# classification file with mono-exonic isoforms retained
class_file <- preFilteredClassFile %>% filter(filter_result == "Isoform")
monoexonic <- class_file[class_file$subcategory == "mono-exon" & class_file$structural_category != "FSM","isoform"]
class_file <- class_file %>% filter(!isoform %in% monoexonic)

# expression
demux <- list.files(path = paste0(rootDir,"/5_cupcake/6_demux"), pattern = "demux_fl_count.csv", full = T)
demux <- fread(demux, data.table = F)
demuxFiltered  <- demux %>% filter(id %in% class_file$isoform) %>% tibble::column_to_rownames(., var = "id")

# reasons why isoforms filtered
ReasonFiltered <- list.files(path = paste0(rootDir,"/5_cupcake/7_sqanti3"), pattern = "adultSortedNuclei_collapsed_filtering_reasons", full = T)
ReasonFiltered  <- fread(ReasonFiltered, data.table = F)

# isolate the reads that are corresponding to the artifacts
artifacts <- preFilteredClassFile[preFilteredClassFile$filter_result == "Artifact","isoform"]
passedReads <- preFilteredClassFile[preFilteredClassFile$filter_result == "Isoform","isoform"]
artifactDemux <- demux[demux$id %in% artifacts,] 
artifactDemux <- merge(artifactDemux, ReasonFiltered[,c("isoform","reasons")], by.x = "id", by.y = "isoform")
artifactSummary <- artifactDemux %>% 
  select(-id) %>%
  group_by(reasons) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# stepwise stats filtering
stage1Stats <- demuxStats[,c("sample","num_seqs")]
stage2Stats <- as.data.frame(mappedStats %>% group_by(sample) %>% tally(name = "aligned_seqs"))
stage3Stats <- as.data.frame(filteredAlignment %>% group_by(sample) %>% tally(name = "corrected_seqs"))
stage4Stats <- as.data.frame(demux %>% select(-id) %>% apply(.,2,sum))       
colnames(stage4Stats) <- "collapsedReads"
passedReads <- preFilteredClassFile[preFilteredClassFile$filter_result == "Isoform","isoform"] 
stage5Stats <- as.data.frame(demux %>% filter(id %in% passedReads) %>% select(-id) %>% apply(.,2,sum))              
colnames(stage5Stats) <- "finalReads"
# demux file to keep only the finalised isoforms
stage6Stats <- as.data.frame(demuxFiltered %>% apply(.,2,sum)) 
colnames(stage6Stats) <- "monoFilteredReads"

# filter demux file
# Define a custom function to count non-zero elements in a row
count_non_zero <- function(row) {
  return(sum(row != 0))
}
nreads <- apply(demuxFiltered,1,sum)
nsample <- apply(demuxFiltered, 1, count_non_zero)
demuxFiltered$nreads <- nreads
demuxFiltered$nsample <- nsample
demuxFiltered <- demuxFiltered %>% filter(nreads >= 2, nsample >= 2)

stage7Stats <- as.data.frame(demuxFiltered %>% select(-nreads, -nsample) %>% apply(.,2,sum)) 
colnames(stage7Stats) <- "expressionFilteredReads"
stage4Stats <- merge(manifest[,c("ID","sample")], stage4Stats, by.x = "ID", by.y = 0)
stage5Stats <- merge(manifest[,c("ID","sample")], stage5Stats, by.x = "ID", by.y = 0)
stage6Stats <- merge(manifest[,c("ID","sample")], stage6Stats, by.x = "ID", by.y = 0)
stage7Stats <- merge(manifest[,c("ID","sample")], stage7Stats, by.x = "ID", by.y = 0)

fullStats <- Reduce(function(x, y) merge(x, y, by = "sample", all = TRUE), 
                    list(stage1Stats, stage2Stats, stage3Stats, stage4Stats, stage5Stats, stage6Stats, stage7Stats)) 

fullStats <- fullStats %>% select(-contains("ID")) %>% 
  mutate(percCorrectedAligned = corrected_seqs/num_seqs * 100, 
         percCollapsed = collapsedReads/num_seqs * 100,
         percFinal = finalReads/num_seqs * 100,
         percMonofiltered = monoFilteredReads/num_seqs * 100,
         percExpressionFiltered = expressionFilteredReads/num_seqs * 100) 

fullStats <- merge(fullStats, manifest, by = "sample")

artifactSummary$reasons <- gsub("High perc_A_downstream_TTS", "intra-priming", artifactSummary$reasons)
artifactSummary$reasons <- gsub("Out RTS_stage", "RT switching", artifactSummary$reasons)
reshape2::melt(artifactSummary, value.name = "removedFL", variable.name = "ID") %>%
  merge(., fullStats, by = "ID") %>% 
  mutate(percRemoved = removedFL/num_seqs * 100) %>% 
  ggplot(., aes(x = reasons, y = percRemoved)) + 
  geom_boxplot(outliers = FALSE) +
  geom_jitter(aes(colour = sample)) + theme_classic() +
  labs(x = "Reasons filtered", y = "Percentage of reads removed")
```

## Expression threshold

To reduce false negatives and increase confidence of identifying "true" isoforms, we can apply an expression filter threshold i.e. an isoform is kept only if it has more than 2 reads across any 2 samples.

The below plot shows the proportion of reads retained after applying a step-wise threshold. Each line corresponds to the minimum number of reads and the x-axis refers to the minimum number of samples. 

```{r sensitivity, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: Line plot of the number of reads retained post SQANTI filtering after applying an expression threshold"}

demuxFiltered  <- demux %>% filter(id %in% class_file$isoform) %>% tibble::column_to_rownames(., var = "id")

sensitivity <- demuxFiltered 
colnames(sensitivity) <- paste0("FL.", colnames(sensitivity))
sensitivtyOutput <- no_of_isoforms_sample(sensitivity)
sensitivtyOutput[[3]]$cate_sumFL <- categories(as.numeric(as.character(sensitivtyOutput[[3]]$sum_FL)))

dat2 <- sensitivtyOutput[[3]] %>%
  group_by(num_samples, cate_sumFL) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = count / sum(count)) %>%
  mutate(cate_sumFL = factor(cate_sumFL, levels = c(seq(1:9),"10","10-20","21-30","31-40","41-50","> 50")))

ggplot(dat2, aes(x = as.factor(num_samples), y = proportion, colour = cate_sumFL, group = cate_sumFL)) +
  geom_point() +
  geom_line() + scale_y_continuous(labels = perc_lab)  + 
  labs(x = "Minimum number of samples", y = "Reads (%)",
       colour = "Minimum number of reads") +
  theme_classic() 
  
```

# QC: Sample reproducibility

We anticipate that while the number of reads may differ across the samples, the extent of reads retained after each processing step will be consistent. Any deviations will suggest a sample/batch failure. 

The plot displays the proportion of reads in each sample **retained** after the following steps:  
    1. Alignment and correction  
    2. Collapsing reads to transcripts  
    3. Filtering reads for RT artifacts  
    4. Removing mono-exonic non-FSM reads  
    5. Applying a minimum expression threshold  

```{r fullStats, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: proportion of reads retained after each processing/filtering stage"}
fullStatsPlot <- fullStats %>% select(sample, contains("perc")) %>% 
  reshape2::melt(variable.name = "stage", value.name = "perc") %>% 
  filter(perc != "NA") 

xaxislevelsF1 <- c("percCorrectedAligned","percCollapsed","percFinal","percMonofiltered", "percExpressionFiltered")
xaxislabelsF1 <- c("Alignment", "Collapse", "SQANTI filtering", "monoexonic filtering", "Expression filtering")
fullStatsPlot$stage = factor(fullStatsPlot$stage,
                             labels = xaxislabelsF1, 
                             levels = xaxislevelsF1,
                             ordered=TRUE)

ggplot(fullStatsPlot , aes(x = stage, y = perc, group = sample, colour = sample)) +
  geom_point() +
  geom_line() +
  theme_classic() + 
  labs(y = "Percentage of reads", x = "Stage") +
  ylim(0,100)
```

# QC: Biological validity

The following sections assess whether the samples sequenced reflect biological phenomena and if there are any covariates that should be accounted in downstream analysis. 

## Principle component analysis

```{r pca, fig.width = 12, fig.height = 6, echo = FALSE}

res <- as.data.frame(t(demuxFiltered)) 
pca_res <- prcomp(as.matrix(res), scale. = TRUE)
pca.scores <- as.data.frame(pca_res$x)
pca_res.pca <- pca_res$sdev^2/sum(pca_res$sdev^2)
pca_res.finalpca <- pca.scores[,which(pca_res.pca > 0.01)]

```

To identify and visually inspect potential outliers we performed principal component analysis on the number of read counts across the isoforms in the final dataset. We identified `r nrow(pca.scores[,which(pca_res.pca > 0.01)])` PCs which explained > 1% of the variance and focused on these for characterisation. 


```{r pca_overallplot, fig.width = 12, fig.height = 6, echo = FALSE}
res <- as.data.frame(t(demuxFiltered)) 
pca_res <- prcomp(as.matrix(res), scale. = TRUE)
pca.scores <- as.data.frame(pca_res$x)
pca_res.pca <- pca_res$sdev^2/sum(pca_res$sdev^2)
pca_res.finalpca <- pca.scores[,which(pca_res.pca > 0.01)]

par(mar = c(4,4,1,1))
pca_var_explained <- (pca_res$sdev^2 / sum(pca_res$sdev^2)) * 100
plot(1:20, pca_var_explained[1:20], type = "b", ylab = "% variance explained", xlab = "Principal Components")
abline(h=1, col = "red")
```

```{r pca_plot, fig.width = 12, fig.height = 4, echo = FALSE}
# Extract PC1 and PC2 scores
pca.scores$cellType <- manifest[match(row.names(pca.scores), manifest$ID), "cellType"]         # Add cellType as a column
pca.scores$sex <- manifest[match(row.names(pca.scores), manifest$ID), "sex"]         # Add cellType as a column
pca.scores$individual <- manifest[match(row.names(pca.scores), manifest$ID), "individual"]         # Add cellType as a column

# List of grouping variables (e.g., factors for coloring)
group_vars <- c("cellType", "sex", "individual")

# List of PC combinations to plot
pc_combinations <- list(
  c("PC1", "PC2"),
  c("PC1", "PC3")
)

# Loop through group variables and PC combinations
PCA_plots <- list()
count=1
for (group_var in group_vars) {
  for (pc_pair in pc_combinations) {
    x_pc <- pc_pair[1]
    y_pc <- pc_pair[2]
    
    # Dynamically generate plot
    p <- ggplot(pca.scores, aes_string(x = x_pc, y = y_pc, color = group_var)) +
      geom_point(size = 3) + 
      labs(
        title = paste0("Coloured by ", group_var),
        subtitle = paste("PCA:", x_pc, "vs", y_pc),
        x = paste("Principal Component", substr(x_pc, 3, 3)),
        y = paste("Principal Component", substr(y_pc, 3, 3))
      ) +
      theme_classic() +
      scale_color_brewer(palette = "Set2")
    
    # Print the plot
    PCA_plots[[count]] <- p
    count = count + 1
  }
}

for (i in seq(1, length(PCA_plots), by = 2)) {
  # Select the next 2 plots (if available)
  current_plots <- PCA_plots[i:min(i + 1, length(PCA_plots))] 
  
  # Print the 2 plots in a grid layout
  print(plot_grid(plotlist = current_plots, ncol = 2))
}

```

## Normalisation

```{r normalisation, fig.width = 12, fig.height = 6, echo = FALSE}

# calculate TPM 
# https://github.com/ConesaLab/SQANTI3/wiki/Running-SQANTI3-Quality-Control
# raw FL read for transcript in sample 1/total FL transcript in sample 1 * 10^6
demuxCollapsed <- demux %>% filter(id %in% preFilteredClassFile$isoform)
row.names(demuxCollapsed) <- demux$id
demuxTPM <- apply(demuxCollapsed %>% select(-id),2,function(x){x/sum(x) * 1000000})
demuxFilteredTPM <- apply(demuxFiltered,2,function(x){x/sum(x) * 1000000})

```

## Gene expression

We expect the gene expression of house keeping genes to be consistent across all samples, and differences in cell-specific genes for sorted populations.

```{r geneCellExpressions, fig.width = 12, fig.height = 6, echo = FALSE}
plot_geneExp <- function(gene){
  gclass_file <- preFilteredClassFile %>% filter(associated_gene == gene) %>% select(isoform, associated_gene)
  
  geneExp <- merge(gclass_file, demuxTPM, by.x = "isoform", by.y = 0) %>% 
    select(-isoform,-associated_gene) %>%
    apply(., 2, sum) %>% 
    reshape2::melt(value.name = "exp") 
  
  geneExp <- merge(manifest, geneExp, by.x = "ID", by.y = 0)
  geneExp <<- geneExp  
  
  if("cellType" %in% colnames(manifest)){
  p <- ggplot(geneExp, aes(x = cellType, y = exp)) + geom_boxplot(outliers = FALSE) + 
    geom_jitter(size = 3, aes(colour = individual)) + 
    theme_classic() + 
    labs(x = "Sorted populations", y = "Expression", title = gene)
  }
  
  return(p)
}

housekeepingGene <- c("GAPDH")
plot_geneExp("GAPDH")

```

### Cell-specific genes

```{r neuronalExpression, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: neuronal-specific gene expressions"}

if("cellType" %in% colnames(manifest)){
  NeuronalGenes <- c("CABP1","PRRG3","RBFOX3","ENO2")
  NeuonralGenePlots <- lapply(NeuronalGenes, function(gene) plot_geneExp(gene))
  plot_grid(plotlist = NeuonralGenePlots)
}else{
  message("FANS not performed; cell-specific gene plots subsequently not displayed")
}
```

```{r oligoCellExpressions, fig.width = 12, fig.height = 6, echo = FALSE, fig.bottomcaption = TRUE, fig.cap="\\raggedright Figure: oligodendrocyte-specific gene expressions"}

OligoGenes <-  c("SOX10", "OPALIN","ST18","OLIG1")
OligoGenePlots <- lapply(OligoGenes, function(gene) plot_geneExp(gene))
plot_grid(plotlist = OligoGenePlots)

```


